Dynamical modeling of tidal streams

### 2013-11-20 tex/ms.tex
streams: more discussion

\renewcommand{\vec}[1]{\ensuremath{\mathbf{#1}}}
\newcommand{\unitvec}[1]{\ensuremath{\mathbf{\hat{#1}}}}
\newcommand{\vecx}{\ensuremath{\vec{x}}}
\newcommand{\vecv}{\ensuremath{\vec{v}}}
\newcommand{\vech}{\ensuremath{\vec{H}}}
\newcommand{\vecj}{\ensuremath{\vec{J}}}
\newcommand{\veco}{\ensuremath{\vec{\Omega}}}
\newcommand{\veca}{\ensuremath{\boldsymbol\theta}}
\newcommand{\df}{\ensuremath{f}}
\newcommand{\tdf}{\ensuremath{\mathrm{tDF}}}
\newcommand{\paramsdiff}{\ensuremath{\vec{p}_\Delta}}
\newcommand{\paramsdf}{\ensuremath{\vec{p}_{\mathrm{DF}}}}
\newcommand{\paramspot}{\ensuremath{\vec{p}_\Phi}}

@@

In action--angle space the dynamics of stream formation is much
simpler. A star received an offset ($\Delta \vecj,\Delta \veca$) at
time $-\Delta t$ from the action--angle coordinates of the progenitor
$(\vecj^p,\veca^p)$, similar to the small offset ($\Delta \vecx,\Delta
\vecv)$ in the real-space description. The offset in the actions
corresponds to an offset in the frequencies $\Delta \veco$;
ultimately, this offset is responsible for the spreading of the stream
over long stretches on the sky. Because both the source's angles and
the star's angles increase linearly with time, albeit at different
frequencies, the difference in the angles also increases linearly in
time. Therefore at time $t=0$ the difference in angles is
\begin{equation}
\Delta \veca(t=0) = \Delta \veca(-\Delta t) + \Delta \veco \Delta t\,,
\end{equation}
while the difference in actions is constant
\begin{equation}
\Delta \vecj(t=0) = \Delta \vecj(-\Delta t)\,.
\end{equation}
As the initial offset $\Delta \veca(-\Delta t$ in angles is small
compared to the growth $\Delta\veco\Delta t$, $\Delta \veca(t=0)
\approx \Delta \veco \Delta t$.

@@

@@

\subsection{Real-space model}\label{sec:realmethod}

@@

A generative model for the positions and velocities
($\vecx_i,\vecv_i$) of stars $i$ in a tidal stream is then fully
specified by: (a) the gravitational potential $\Phi$, (b) the current
phase-space position of the source $(\vecx_s,\vecv_s)(t=0)$, which can
be integrated backwards in time by reversing the Hamiltonian flow
defined by $\Phi$, (c) the time $\Delta t_i$ at which each star was
removed from the source, and (d) the distribution $f(\Delta
\vecx,\Delta \vecv|\paramsdiff)$, characterized by some parameters
$\paramsdiff$, of initial phase-space offsets from which the actual
offsets $(\Delta \vecx_i,\Delta \vecv_i)$ are drawn. In practice, $f$
can be taken to be a Gaussian, characterized by a width in position
$\sigma_x$ and in velocity $\sigma_v$, corresponding to the tidal
radius and escape velocity of the progenitor, respectively. However,
more complicated models for the initial phase-space distribution of
the debris can be used here as well (see discussion below). In the
case of a Gaussian $f$, the source can be thought of as the mean of
this Gaussian, with the additional requirement that this mean is an
orbit in the potential $\Phi$.

The probability for observing a star today at $(\vecx,\vecv)$ given
items (a) through (d) is then
\begin{align}\label{eq:realdft}
  p(\vecx_i,\vecv_i | \Phi,\vecx_s,\vecv_s,\Delta t_i,\paramsdiff) 
  & = p(\vecx_i(-\Delta t),\vecv_i(-\Delta t) | \Phi,\vecx_s,\vecv_s,\Delta t_i,\paramsdiff) \,,\nonumber\\
  & = p([\vecx_i-\vecx_s](-\Delta t),[\vecv_i-\vecv_s](-\Delta t) | \Phi,\vecx_s,\vecv_s,\Delta t_i,\paramsdiff) \,\nonumber\\
  & = p(\Delta\vecx_i(-\Delta t),\Delta\vecv_i(-\Delta t) | \Phi,\vecx_s,\vecv_s,\Delta t_i,\paramsdiff)\,,\\
  & = f(\Delta\vecx_i(-\Delta t),\Delta\vecv_i(-\Delta t) | \Phi,\vecx_s,\vecv_s,\Delta t_i,\paramsdiff)\, ,\nonumber
\end{align}
where the first equality holds because the Hamiltonian flow is unique
and conserves phase-space volume by Liousville's theorem and the
second one because the Jacobian of the transformation
$(\vecx_i,\vecv_i)(-\Delta t) \rightarrow (\vecx_i,\vecv_i)(-\Delta t)
- (\vecx_s,\vecv_s)(-\Delta t)$ is one. The final probability can be
evaluated using the parameters $\paramsdiff$ following (d).

The time $\Delta t_i$ at which a star was removed is unobservable and
the probability in \eqnname~(\ref{eq:realdft}) should therefore be
integrated over this time using a prior $p(\Delta t_i|\ldots)$, where
$\ldots$ indicates that this prior could depend on other factors such
as the potential and the progenitor orbit\footnote{For example, one
  could insist that stars are removed around pericenter passages. Note
  that this assumption would require the source orbit to be the actual
  progenitor orbit.}. For simplicity, we will assume a uniform prior
for the times $\Delta t$. The probability for a single star then becomes
\begin{align}\label{eq:realdf}
  p(\vecx_i,\vecv_i | \Phi,\vecx_s,\vecv_s,\paramsdiff) & = \int \dd
  \Delta t_i \,p(\vecx_i,\vecv_i | \Phi,\vecx_s,\vecv_s,\Delta t_i,
  \paramsdiff)\,.
\end{align}
If the progenitor's current phase-space position is unknown, then this
probability should be further marginalized over $(\vecx_s,\vecv_s)$;
for multiple stars, this marginalization is done after multiplying the
individual-star likelihoods in \eqnname~(\ref{eq:realdf}).

In practice, the likelihood in \eqnname~(\ref{eq:realdf}) can be
calculated as follows: for a given potential $\Phi$ and source
$(\vecx_s,\vecv_s)$, we (a) integrate the orbits of all stars
$(\vecx_i,\vecv_i)$ and (b) that of the source backward in time for
many Gyr and record the phase-space positions over time on a grid $j$
in $\Delta^j_i$; (c) we evaluate $f(\Delta \vecx_i,\Delta
\vecv_i)(\Delta t^j_i)$ for all stars $i$ and times $j$; (d) $f(\Delta
\vecx_i,\Delta \vecv_i)(\Delta t^j_i)$ is then summed over $j$ and the
logarithm of that is summed over $i$. The most computationally
expensive part of this calculation is the integration of the various
orbits. The evaluation of the likelihood for different parameters
\paramsdiff\ of $f$ can be efficiently computed by caching the orbit
integrations of (a) and (b); different source orbits
$(\vecx_s,\vecv_s)$ can also make use of the stored orbits in (a).

@@

In the action-angle formalism, a star is removed from the source with
a small offset $(\Delta \vecj,\Delta \veca)(-\Delta t)$ from the
source's actions $\vecj_s$ and angles $\veca_s$ a time $\Delta t$ ago,
similar to the small offset $(\Delta \vecx,\Delta \vecv)(-\Delta t)$
in the real-space model. The offset in the actions corresponds to an
offset in the frequencies $\Delta \veco$; ultimately, this offset is
responsible for the spreading of the stream over long stretches on the
sky. Because both the source's angles and the star's angles increase
linearly with time, albeit at different frequencies, the difference in
the angles also increases linearly in time. Therefore at time $t=0$
the difference in angles is
\begin{equation}
\Delta \veca(t=0) = \Delta \veca(-\Delta t) + \Delta \veco \Delta t\,,
\end{equation}
while the difference in actions is constant
\begin{equation}
\Delta \vecj(t=0) = \Delta \vecj(-\Delta t)\,.
\end{equation}

With these preliminaries, we can write down a generative model for the
positions and velocities ($\vecx_i,\vecv_i$) of stars $i$ in a tidal
stream in the action-angle formalism by specifying: (a) the
gravitational potential $\Phi$, (b) the current action-angle
coordinates of the source $(\vecj_s,\veca_s)(t=0)$, (c) the time
$\Delta t_i$ at which each star was removed from the source, and (d)
the distribution $f(\Delta \vecj,\Delta \veca|\paramsdiff)$,
characterized by some parameters $\paramsdiff$, of initial phase-space
offsets from which the actual offsets $(\Delta \vecj_i,\Delta
\veca_i)$ are drawn. In practice, $f$ can again be taken to be a
Gaussian, although more general models could take the expected
`bowtie' structure of the action distribution into account
\citep{Eyre11a}.

The probability for observing a star today at $(\vecx,\vecv)$ given
items (a) through (d) is then
\begin{align}\label{eq:aadft}
  p(\vecx_i,\vecv_i | \Phi,\vecj_s,\veca_s,\Delta t_i,\paramsdiff) 
  & = p(\vecj_i,\veca_i | \Phi,\vecj_s,\veca_s,\Delta t_i,\paramsdiff) \,,\nonumber\\
  & = p(\vecj_i-\vecj_s,\veca_i-\veca_s-\Delta \veco_i\Delta t_i | \Phi,\vecj_s,\veca_s,\Delta t_i,\paramsdiff) \,,\\
  & = f(\Delta \vecj_i,\Delta \veca_i(-\Delta t_i) | \Phi,\vecj_s,\veca_s,\Delta t_i,\paramsdiff) \,,\nonumber
\end{align}
where the first equality holds because the Jacobian of the
$(\vecx,\vecv)\rightarrow(\vecj,\veca)$ transformation is unity and
the second equality is a simple linear transformation. The final
probability can again be evaluated using the parameters
\paramsdiff\ following (d).

>>>>

\renewcommand{\vec}[1]{\ensuremath{\mathbf{#1}}}
\newcommand{\unitvec}[1]{\ensuremath{\mathbf{\hat{#1}}}}
\newcommand{\vecx}{\ensuremath{\vec{x}}}
\newcommand{\vecv}{\ensuremath{\vec{v}}}
\newcommand{\vech}{\ensuremath{\vec{H}}}
\newcommand{\vecj}{\ensuremath{\vec{J}}}
\newcommand{\vecp}{\ensuremath{\vec{p}}}
\newcommand{\veco}{\ensuremath{\vec{\Omega}}}
\newcommand{\veca}{\ensuremath{\boldsymbol\theta}}
\newcommand{\df}{\ensuremath{f}}
\newcommand{\tdf}{\ensuremath{\mathrm{tDF}}}
\newcommand{\paramsdiff}{\ensuremath{\vec{p}_\Delta}}
\newcommand{\paramsdf}{\ensuremath{\vec{p}_{\mathrm{DF}}}}
\newcommand{\paramspot}{\ensuremath{\vec{p}_\Phi}}

@@

In action--angle space the dynamics of stream formation is much
simpler. A star received an offset ($\Delta \vecj,\Delta \veca$) at
time $-\Delta t$ from the action--angle coordinates of the progenitor
$(\vecj^p,\veca^p)$, similar to the small offset ($\Delta \vecx,\Delta
\vecv)$ in the real--space description. The offset in the actions
corresponds to an offset in the frequencies $\Delta \veco$;
ultimately, this offset is responsible for the spreading of the stream
over long stretches on the sky. Because both the progenitor's angles and
the star's angles increase linearly with time, albeit at different
frequencies, the difference in the angles also increases linearly in
time. Therefore at time $t=0$ the difference in angles is
\begin{equation}
\Delta \veca(t=0) = \Delta \veca(-\Delta t) + \Delta \veco \Delta t\,,
\end{equation}
while the difference in actions is constant
\begin{equation}
\Delta \vecj(t=0) = \Delta \vecj(-\Delta t)\,.
\end{equation}
As the initial offset $\Delta \veca(-\Delta t$ in angles is small
compared to the growth $\Delta\veco\Delta t$, $\Delta \veca(t=0)
\approx \Delta \veco \Delta t$.

@@

\section{Generative models of tidal streams}\label{sec:method}

The discussions of the dynamics of stream formation in the previous
section directly give rise of a generative model of a tidal stream. A
star in a tidal stream is \emph{generated} through being stripped a
time $\Delta t$ ago from a progenitor $(\vecx^p,\vecv^p) \equiv
(\vecj^p,\veca^p)$ by receiving a small offset $(\Delta \vecx,\Delta
\vecv) \equiv (\Delta \vecj,\Delta \veca)$. The offset then evolves
using the dynamical equations given in the previous section and the
star is observed at time $t=0$ (\ie, today). Because phase-space
density is conserved by Liousville's theorem, a probabilistic model
results by relating the likelihood $p(\vecx,\vecv|\vecp)$ (where
$\vecp$ denotes the set of all model parameters; see below) for the
observed phase-space position today to the distribution of initial
offsets. This probabilistic model can be marginalized over missing
observational data or extended to include a model for outliers.

\subsection{Real--space modeling}

In more detail, in the real--space approach the model can be specified
by: (a) the gravitational potential $\Phi$, (b) the current
phase-space position of the progenitor $(\vecx_p,\vecv_p)(t=0)$, which
can be integrated backwards in time by reversing the Hamiltonian flow
defined by $\Phi$, (c) the time $\Delta t_i$ at which each star was
removed from the progenitor, and (d) the distribution $f(\Delta
\vecx,\Delta \vecv|\paramsdiff)$, characterized by some parameters
$\paramsdiff$, of initial phase-space offsets from which the actual
offsets $(\Delta \vecx_i,\Delta \vecv_i)$ are drawn. In practice, $f$
can be taken to be a Gaussian, characterized by a width in position
$\sigma_x$ and in velocity $\sigma_v$, related to the tidal radius and
escape velocity of the progenitor, respectively. However, more
complicated models for the initial phase-space distribution of the
debris can be used here as well.

The probability for observing a star today at $(\vecx,\vecv)$ given
items (a) through (d) is then
\begin{align}\label{eq:realdft}
  p(\vecx_i,\vecv_i | \Phi,\vecx_p,\vecv_p,\Delta t_i,\paramsdiff) 
  & = p(\vecx_i(-\Delta t),\vecv_i(-\Delta t) | \Phi,\vecx_p,\vecv_p,\Delta t_i,\paramsdiff) \,,\nonumber\\
  & = p([\vecx_i-\vecx_p](-\Delta t),[\vecv_i-\vecv_p](-\Delta t) | \Phi,\vecx_p,\vecv_p,\Delta t_i,\paramsdiff) \,\nonumber\\
  & = p(\Delta\vecx_i(-\Delta t),\Delta\vecv_i(-\Delta t) | \Phi,\vecx_p,\vecv_p,\Delta t_i,\paramsdiff)\,,\\
  & = f(\Delta\vecx_i(-\Delta t),\Delta\vecv_i(-\Delta t) | \Phi,\vecx_p,\vecv_p,\Delta t_i,\paramsdiff)\, ,\nonumber
\end{align}
where the first equality holds because the Hamiltonian flow is unique
and conserves phase-space volume by Liousville's theorem and the
second one because the Jacobian of the transformation
$(\vecx_i,\vecv_i)(-\Delta t) \rightarrow (\vecx_i,\vecv_i)(-\Delta t)
- (\vecx_p,\vecv_p)(-\Delta t)$ is one. The final probability can be
evaluated using the parameters $\paramsdiff$ following (d).

The time $\Delta t_i$ at which a star was removed is unobservable and
the probability in \eqnname~(\ref{eq:realdft}) should therefore be
integrated over this time using a prior $p(\Delta t_i|\ldots)$, where
$\ldots$ indicates that this prior could depend on other factors such
as the potential and the progenitor orbit\footnote{For example, one
  could insist that stars are removed around pericenter
  passages.}. For simplicity, we can assume a uniform prior for the
times $\Delta t$. The probability for a single star then becomes
\begin{align}\label{eq:realdf}
  p(\vecx_i,\vecv_i | \Phi,\vecx_p,\vecv_p,\paramsdiff) & = \int \dd
  \Delta t_i \,p(\vecx_i,\vecv_i | \Phi,\vecx_p,\vecv_p,\Delta t_i,
  \paramsdiff)\,.
\end{align}
If the progenitor's current phase-space position is unknown, then this
probability should be further marginalized over $(\vecx_p,\vecv_p)$;
for multiple stars, this marginalization is done after multiplying the
individual-star likelihoods in \eqnname~(\ref{eq:realdf}).

In practice, the likelihood in \eqnname~(\ref{eq:realdf}) can be
calculated as follows: for a given potential $\Phi$ and progenitor
$(\vecx_p,\vecv_p)$, we (a) integrate the orbits of all stars
$(\vecx_i,\vecv_i)$ and (b) that of the progenitor backward in time for
many Gyr and record the phase-space positions over time on a grid $j$
in $\Delta^j_i$; (c) we evaluate $f(\Delta \vecx_i,\Delta
\vecv_i)(\Delta t^j_i)$ for all stars $i$ and times $j$; (d) $f(\Delta
\vecx_i,\Delta \vecv_i)(\Delta t^j_i)$ is then summed over $j$ and the
logarithm of that is summed over $i$. The most computationally
expensive part of this calculation is the integration of the various
orbits. The evaluation of the likelihood for different parameters
\paramsdiff\ of $f$ can be efficiently computed by caching the orbit
integrations of (a) and (b); different progenitor orbits
$(\vecx_p,\vecv_p)$ can also make use of the stored orbits in (a).

The approach outlined in this section is straightforward, but very
computationally expensive, as it requires large numbers of orbit
integrations to explore all plausible removal times $\Delta t$,
progenitor orbits, and potentials. Moreover, when the data have large
uncertainties, more orbit integrations are necessary to explore each
data points uncertainty range. While it might be possible to speed up
the computation through linear approximations for nearby orbits, the
action--angle formalism allows for significant and elegant
simplifications and I consider the real--space approach further.

\subsection{Action--angle stream modeling}

Similar as for the real--space approach, we can write down a
generative model for the positions and velocities ($\vecx_i,\vecv_i$)
of stars $i$ in a tidal stream in the action--angle formalism by
specifying: (a) the gravitational potential $\Phi$, (b) the current
action-angle coordinates of the progenitor $(\vecj_p,\veca_p)(t=0)$,
(c) the time $\Delta t_i$ at which each star was removed from the
progenitor, and (d) the distribution $f(\Delta \vecj,\Delta
\veca|\paramsdiff)$, characterized by some parameters $\paramsdiff$,
of initial phase-space offsets from which the actual offsets $(\Delta
\vecj_i,\Delta \veca_i)$ are drawn. In practice, $f$ can again be
taken to be a Gaussian, although more general models could take the
expected `bowtie' structure of the action distribution into account
(\citealt{Eyre11a}; see also \figurename~\ref{fig:gd1_jt}).

The probability for observing a star today at $(\vecx,\vecv)$ given
items (a) through (d) is then
\begin{align}\label{eq:aadft}
  p(\vecx_i,\vecv_i | \Phi,\vecj_p,\veca_p,\Delta t_i,\paramsdiff) 
  & = p(\vecj_i,\veca_i | \Phi,\vecj_p,\veca_p,\Delta t_i,\paramsdiff) \,,\nonumber\\
  & = p(\vecj_i-\vecj_p,\veca_i-\veca_p-\Delta \veco_i\Delta t_i | \Phi,\vecj_p,\veca_p,\Delta t_i,\paramsdiff) \,,\\
  & = f(\Delta \vecj_i,\Delta \veca_i(-\Delta t_i) | \Phi,\vecj_p,\veca_p,\Delta t_i,\paramsdiff) \,,\nonumber
\end{align}
where the first equality holds because the Jacobian of the
$(\vecx,\vecv)\rightarrow(\vecj,\veca)$ transformation is unity and
the second equality is a simple linear transformation. The final
probability can again be evaluated using the parameters
\paramsdiff\ following (d).

As in the real--space formalism, we should integrate
\eqnname~(\ref{eq:aadft}) over the time $\Delta t_i$ at which the star
is removed from the progenitor (cf.~\eqnname~[\ref{eq:realdf}]):
\begin{equation}\label{eq:aadf}
  p(\vecx_i,\vecv_i | \Phi,\vecj_p,\veca_p,\paramsdiff) = \int \dd
  \Delta t_i\, f(\Delta \vecj_i,\Delta \veca_i(-\Delta t_i) |
  \Phi,\vecj_p,\veca_p,\Delta t_i,\paramsdiff) \,.
\end{equation}
Without further assumptions the action-angle formalism proceeds in the
same way as the real--space formalism, evaluating
\eqnname~(\ref{eq:aadft}) for a grid $j$ of times $\Delta t_i^j$ and
summing $f(\Delta \vecj_i,\Delta \veca_i(-\Delta t^j_i) |
\Phi,\vecj_p,\veca_p,\Delta t_i,\paramsdiff)$ over $j$.

Even without further assumptions about the shape of $f(\Delta
\vecj,\Delta \veca|\paramsdiff)$, the action-angle formalism already
allows for a simplification over the real--space method. Because the
progenitor orbit is specified in action-angle coordinates, it can be
changed \emph{without} expensive orbit integrations of all possible
source orbits. As the progenitor's current phase-space position is a
six-dimensional parameter, this is a major simplification of the
dynamical model when the source orbit is unknown. Note that this
simplification does require us to be able to compute $\veco(\vecj)$,
to calculate the new source frequencies $\veco_p$ when changing the
source actions $\vecj_p$. This mapping can be constructed using a
small number of (action,frequency) calculations covering the range of
possible source actions and is relatively easy to construct due to the
approximately one-dimensional structure of a stream (see discussion at
the end of \sectionname~\ref{sec:dynamics}).

We can simplify \eqnname~(\ref{eq:aadf}) by assuming (a) that
$f(\Delta \vecj,\Delta \veca|\paramsdiff)$ is separable into
$f_{\veca}(\Delta \veca|\paramsdiff)\,f_{\vecj}(\Delta
\vecj|\paramsdiff)$ and (b) that the distribution $f_{\veca}(\Delta
\veca|\paramsdiff)$ is Gaussian. In what follows, we will assume that
this Gaussian is isotropic with dispersion $\sigma_\theta$, but all of
the results are the same after appropriate diagonalization and
rescaling of the angles for more general covariance matrices. In this
case, \eqnname~(\ref{eq:aadf}) can be written as
\begin{equation}
\begin{split}
  p(\vecx_i,\vecv_i & | \Phi,\vecj_p,\veca_p,\paramsdiff) 
   = \\
   & f_{\vecj}(\Delta \vecj_i|\paramsdiff) \int \dd \Delta t_i\,
  \frac{1}{(2\,\pi)^{3/2}\,\sigma_\theta^3}\,\exp\left(-\frac{1}{2\,\sigma_\theta^2}\left[\sum_{R,\phi,Z}\left(\Delta \veca_i-\Delta \veco_i\Delta t_i\right)^2\right]\right)\,,
\end{split}
\end{equation}
where the sum in the argument of the exponential is over the three
components of $\Delta \veca_i$ and $\Delta \veco_i$. This integral can
be done analytically, resulting in
\begin{equation}\label{eq:aadfsimple}
\begin{split}
  p(\vecx_i,\vecv_i & | \Phi,\vecj_p,\veca_p,\paramsdiff) 
   = \\
   & f_{\vecj}(\Delta \vecj_i|\paramsdiff) \,  \frac{1}{2\,\pi\,\sigma_\theta^2\,\sqrt{\sum_{R,\phi,Z}\Delta \veco_i^2}}\,\exp\left(-\frac{1}{2\,\sigma_\theta^2}\left[\sum_{R,\phi,Z}\Delta \veca^2_i-\frac{\left(\sum_{R,\phi,Z} \Delta \veco_i\Delta \veca_i\right)^2}{\sum_{R,\phi,Z}\Delta \veco_i^2}\right]\right)\,.
\end{split}
\end{equation}
In this expression, $\Delta X^2 \equiv (X_1-X_2)^2$. The expression in
square brackets is positive by using the Cauchy-Schwarz inequality.

A detail that is ignored in \equationname~(\ref{eq:aadfsimple}) is
that the uniform prior on $\Delta t$ should only have support for
$\Delta t > 0$, \ie, for times in the past, which introduces an
additional factor of $[1+\mathrm{erf}(t_0)]$, with
\[ t_0 = \frac{\sqrt{\sum_{R,\phi,Z}\Delta
  \veco_i^2}}{\sqrt{2}\,\sigma_\theta}\,\frac{\sum_{R,\phi,Z} \Delta
  \veco_i\Delta \veca_i}{\sum_{R,\phi,Z}\Delta \veco_i^2}\,.\] Because
the second factor on the right-hand side is the best guess for $\Delta
t_i$, if this is large and negative (\ie, the star was stripped
\emph{in the future}), the probability of the star being stripped from
the progenitor is strongly suppressed. Only when it appears that the
star was only recently stripped with $\Delta t_i$ negative, but
$\approx \sigma_\theta/\sqrt{\sum_{R,\phi,Z}\Delta \veco_i^2}$, is the
suppression for negative stripping times less.

If the stream is one-dimensional in angle space then the time since
the removal from the progenitor becomes unobservable if the progenitor
orbit is unknown: above, determining the time depended on explaining
the three-dimensional angle difference $\Delta \veca$ with the
three-dimensional frequency difference $\Delta \veco$ using a single
time difference; if two of the angle and frequency differences are
unable to be resolved, then the frequency difference and the time
difference become degenerate (aka, you can draw a straight line
connecting any two points). To make use of this fact to simplify the
dynamical model, we write the frequency difference as
\begin{equation}
  \Delta \veco = \Delta \Omega_1\,\unitvec{e}_1\,,
\end{equation}
where $\unitvec{e}_1$ is a unit vector. Then we write the angle
difference $\Delta \veca$ 
\begin{equation}
  \Delta \veca = \Delta \theta_1\,\unitvec{e}_1 + \Delta \veca_\perp\,\unitvec{e}_1^\perp\,,
\end{equation}
where$\unitvec{e}_1^\perp$ is the orthogonal complement of
$\unitvec{e}_1$ (the two directions perpendicular to $\unitvec{e}_1$)
and $\Delta \veca_\perp$ are the components of $\Delta \veca$ in that
direction. We can then write \eqnname~(\ref{eq:aadfsimple}) as
\begin{equation}\label{eq:aadfsimpler}
\begin{split}
  p(\vecx_i,\vecv_i & | \Phi,\vecj_p,\veca_p,\paramsdiff) 
   = \\
   & f_{\vecj}(\Delta \vecj_i|\paramsdiff) \,  \frac{1}{2\,\pi\,\sigma_\theta^2\,\sqrt{\sum_{R,\phi,Z}\Delta \veco_i^2}}\,\exp\left(-\frac{|\Delta\veca_{i,\perp}|^2}{2\,\sigma_\theta^2}\right)\,.
\end{split}
\end{equation}
\Eqnname~(\ref{eq:aadfsimpler}) is equivalent to
\eqnname~(\ref{eq:aadfsimple}), but it expresses the dynamical model
in a slightly different way. Whereas the angle part of
\eqnname~(\ref{eq:aadfsimple}) allows the correct potential and
progenitor to be inferred by asking whether a single time difference
explains the three angle differences given the three frequencies,
\eqnname~(\ref{eq:aadfsimpler}) `rewinds' the star's orbit along the
direction of $\Delta \veco$ and asks whether the angle difference
perpendicular to $\Delta \veco$ is consistent with the initial angle
spread (because it has not evolved since the removal of the star). In
re-writing this, we have not yet used the approximation that the
stream only forms along a single direction in $\Delta \veco$.

@@

\subsection{Real--space model}\label{sec:realmethod}

@@

A generative model for the positions and velocities
($\vecx_i,\vecv_i$) of stars $i$ in a tidal stream is then fully
specified by: (a) the gravitational potential $\Phi$, (b) the current
phase-space position of the source $(\vecx_p,\vecv_p)(t=0)$, which can
be integrated backwards in time by reversing the Hamiltonian flow
defined by $\Phi$, (c) the time $\Delta t_i$ at which each star was
removed from the source, and (d) the distribution $f(\Delta
\vecx,\Delta \vecv|\paramsdiff)$, characterized by some parameters
$\paramsdiff$, of initial phase-space offsets from which the actual
offsets $(\Delta \vecx_i,\Delta \vecv_i)$ are drawn. In practice, $f$
can be taken to be a Gaussian, characterized by a width in position
$\sigma_x$ and in velocity $\sigma_v$, corresponding to the tidal
radius and escape velocity of the progenitor, respectively. However,
more complicated models for the initial phase-space distribution of
the debris can be used here as well (see discussion below). In the
case of a Gaussian $f$, the source can be thought of as the mean of
this Gaussian, with the additional requirement that this mean is an
orbit in the potential $\Phi$.

The probability for observing a star today at $(\vecx,\vecv)$ given
items (a) through (d) is then
\begin{align}\label{eq:realdft}
  p(\vecx_i,\vecv_i | \Phi,\vecx_p,\vecv_p,\Delta t_i,\paramsdiff) 
  & = p(\vecx_i(-\Delta t),\vecv_i(-\Delta t) | \Phi,\vecx_p,\vecv_p,\Delta t_i,\paramsdiff) \,,\nonumber\\
  & = p([\vecx_i-\vecx_p](-\Delta t),[\vecv_i-\vecv_p](-\Delta t) | \Phi,\vecx_p,\vecv_p,\Delta t_i,\paramsdiff) \,\nonumber\\
  & = p(\Delta\vecx_i(-\Delta t),\Delta\vecv_i(-\Delta t) | \Phi,\vecx_p,\vecv_p,\Delta t_i,\paramsdiff)\,,\\
  & = f(\Delta\vecx_i(-\Delta t),\Delta\vecv_i(-\Delta t) | \Phi,\vecx_p,\vecv_p,\Delta t_i,\paramsdiff)\, ,\nonumber
\end{align}
where the first equality holds because the Hamiltonian flow is unique
and conserves phase-space volume by Liousville's theorem and the
second one because the Jacobian of the transformation
$(\vecx_i,\vecv_i)(-\Delta t) \rightarrow (\vecx_i,\vecv_i)(-\Delta t)
- (\vecx_p,\vecv_p)(-\Delta t)$ is one. The final probability can be
evaluated using the parameters $\paramsdiff$ following (d).

The time $\Delta t_i$ at which a star was removed is unobservable and
the probability in \eqnname~(\ref{eq:realdft}) should therefore be
integrated over this time using a prior $p(\Delta t_i|\ldots)$, where
$\ldots$ indicates that this prior could depend on other factors such
as the potential and the progenitor orbit\footnote{For example, one
  could insist that stars are removed around pericenter passages. Note
  that this assumption would require the source orbit to be the actual
  progenitor orbit.}. For simplicity, we will assume a uniform prior
for the times $\Delta t$. The probability for a single star then becomes
\begin{align}\label{eq:realdf}
  p(\vecx_i,\vecv_i | \Phi,\vecx_p,\vecv_p,\paramsdiff) & = \int \dd
  \Delta t_i \,p(\vecx_i,\vecv_i | \Phi,\vecx_p,\vecv_p,\Delta t_i,
  \paramsdiff)\,.
\end{align}
If the progenitor's current phase-space position is unknown, then this
probability should be further marginalized over $(\vecx_p,\vecv_p)$;
for multiple stars, this marginalization is done after multiplying the
individual-star likelihoods in \eqnname~(\ref{eq:realdf}).

In practice, the likelihood in \eqnname~(\ref{eq:realdf}) can be
calculated as follows: for a given potential $\Phi$ and source
$(\vecx_p,\vecv_p)$, we (a) integrate the orbits of all stars
$(\vecx_i,\vecv_i)$ and (b) that of the source backward in time for
many Gyr and record the phase-space positions over time on a grid $j$
in $\Delta^j_i$; (c) we evaluate $f(\Delta \vecx_i,\Delta
\vecv_i)(\Delta t^j_i)$ for all stars $i$ and times $j$; (d) $f(\Delta
\vecx_i,\Delta \vecv_i)(\Delta t^j_i)$ is then summed over $j$ and the
logarithm of that is summed over $i$. The most computationally
expensive part of this calculation is the integration of the various
orbits. The evaluation of the likelihood for different parameters
\paramsdiff\ of $f$ can be efficiently computed by caching the orbit
integrations of (a) and (b); different source orbits
$(\vecx_p,\vecv_p)$ can also make use of the stored orbits in (a).

@@

In the action-angle formalism, a star is removed from the source with
a small offset $(\Delta \vecj,\Delta \veca)(-\Delta t)$ from the
source's actions $\vecj_p$ and angles $\veca_p$ a time $\Delta t$ ago,
similar to the small offset $(\Delta \vecx,\Delta \vecv)(-\Delta t)$
in the real--space model. The offset in the actions corresponds to an
offset in the frequencies $\Delta \veco$; ultimately, this offset is
responsible for the spreading of the stream over long stretches on the
sky. Because both the source's angles and the star's angles increase
linearly with time, albeit at different frequencies, the difference in
the angles also increases linearly in time. Therefore at time $t=0$
the difference in angles is
\begin{equation}
\Delta \veca(t=0) = \Delta \veca(-\Delta t) + \Delta \veco \Delta t\,,
\end{equation}
while the difference in actions is constant
\begin{equation}
\Delta \vecj(t=0) = \Delta \vecj(-\Delta t)\,.
\end{equation}

With these preliminaries, we can write down a generative model for the
positions and velocities ($\vecx_i,\vecv_i$) of stars $i$ in a tidal
stream in the action-angle formalism by specifying: (a) the
gravitational potential $\Phi$, (b) the current action-angle
coordinates of the source $(\vecj_p,\veca_p)(t=0)$, (c) the time
$\Delta t_i$ at which each star was removed from the source, and (d)
the distribution $f(\Delta \vecj,\Delta \veca|\paramsdiff)$,
characterized by some parameters $\paramsdiff$, of initial phase-space
offsets from which the actual offsets $(\Delta \vecj_i,\Delta
\veca_i)$ are drawn. In practice, $f$ can again be taken to be a
Gaussian, although more general models could take the expected
`bowtie' structure of the action distribution into account
\citep{Eyre11a}.

The probability for observing a star today at $(\vecx,\vecv)$ given
items (a) through (d) is then
\begin{align}\label{eq:aadft}
  p(\vecx_i,\vecv_i | \Phi,\vecj_p,\veca_p,\Delta t_i,\paramsdiff) 
  & = p(\vecj_i,\veca_i | \Phi,\vecj_p,\veca_p,\Delta t_i,\paramsdiff) \,,\nonumber\\
  & = p(\vecj_i-\vecj_p,\veca_i-\veca_p-\Delta \veco_i\Delta t_i | \Phi,\vecj_p,\veca_p,\Delta t_i,\paramsdiff) \,,\\
  & = f(\Delta \vecj_i,\Delta \veca_i(-\Delta t_i) | \Phi,\vecj_p,\veca_p,\Delta t_i,\paramsdiff) \,,\nonumber
\end{align}
where the first equality holds because the Jacobian of the
$(\vecx,\vecv)\rightarrow(\vecj,\veca)$ transformation is unity and
the second equality is a simple linear transformation. The final
probability can again be evaluated using the parameters
\paramsdiff\ following (d).

As in the real--space formalism, we should integrate
\eqnname~(\ref{eq:aadft}) over the time $\Delta t_i$ at which the star
is removed from the source (cf.~\eqnname~[\ref{eq:realdf}]):
\begin{equation}\label{eq:aadf}
  p(\vecx_i,\vecv_i | \Phi,\vecj_p,\veca_p,\paramsdiff) = \int \dd
  \Delta t_i\, f(\Delta \vecj_i,\Delta \veca_i(-\Delta t_i) |
  \Phi,\vecj_p,\veca_p,\Delta t_i,\paramsdiff) \,.
\end{equation}
Without further assumptions the action-angle formalism proceeds in the
same way as the real--space formalism, evaluating
\eqnname~(\ref{eq:aadft}) for a grid $j$ of times $\Delta t_i^j$ and
summing $f(\Delta \vecj_i,\Delta \veca_i(-\Delta t^j_i) |
\Phi,\vecj_p,\veca_p,\Delta t_i,\paramsdiff)$ over $j$.
